---
title: "GarnerChildersMuhimpundu_FinalProject"
author: "Seth Garner, Gavin Childers, Julie Muhimpundu"
date: "2026-04-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = "..")
library(tidyverse)
```

```{r}
sleep_df <- read_csv("sleep_data.csv")
sleep_df
```

# Heading 1: Probability Distributions

## 1.1) Qualitative Data

Sleep Disorder
Success == "None"
Fail == "Sleep Apnea" or "Insomnia"

```{r}
# Population proportion (p)
p <- mean(sleep_df$`Sleep Disorder` == "None", na.rm = TRUE)
p

n <- 50

# Mean (mu)
mu <- n * p

# Standard deviation (sigma)
sigma <- sqrt(n * p * (1 - p))

mu
sigma
```
The random variable X, representing the number of individuals with no sleep disorder in a sample of size 50, follows a binomial distribution because:
i) The Number of Trials are Fixed (n = 50)
ii) Each trial has two outcomes (Success, Fail)
iii) The probability of success (p) is constant
iv) Observations are Independent

```{r}
np <- n * p
nq <- n * (1 - p)
np
nq

z <- (mu + 0.5 - mu) / sigma
z

prob <- 1 - pnorm(z)
prob
```
Since np & n(1-p) > 10, the binomial distribution is approximately normal. A continuity correction of 0.5 is applied when converting from the discrete binomial distribution to the continuous normal distribution. The resulting probability represents the likelihood that the number of individuals with no sleep disorder in a sample of 50 exceeds the expected value.

```{r}
# Population proportion
p <- mean(sleep_df$`Sleep Disorder` == "None", na.rm = TRUE)

# Sample size
n <- 50

# Mean of sampling distribution
phat_mean <- p

# Standard error of sampling distribution
phat_se <- sqrt(p * (1 - p) / n)

phat_mean
phat_se
```
The mean of the sampling distribution equals the population proportion p, confirming that $\hat{p} $is an unbiased estimator. The standard error measures the typical variation of the sample proportion from sample to sample, and it decreases as sample size increases.

## 1.2) Quanitative Data

Sleep Duration

The quantitative variable selected is `Sleep Duration`, which measures the number of hours of sleep per individual.

```{r}
ggplot(sleep_df, aes(x = `Sleep Duration`)) +
  geom_histogram(bins = 30, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Sleep Duration",
       x = "Sleep Duration",
       y = "Frequency")
```
The histogram of Sleep Duration does not show a clear bell-shaped or symmetric distribution. Instead, the data appears somewhat unevenly distributed, with mild clustering and no strong single central peak. This suggests that the variable is not approximately normally distributed.

Because the shape is not clearly symmetric and does not follow the typical Gaussian pattern, we conclude that the population distribution of Sleep Duration is not normal.

The sampling distribution of the sample mean can still be approximately normal due to the Central Limit Theorem.


Sample Size: n = 50

X = Sleep Duration
$\bar{X}$ = Sample Mean of Sleep Duration
n = 50

```{r}
mu <- mean(sleep_df$`Sleep Duration`, na.rm = TRUE)
sigma <- sd(sleep_df$`Sleep Duration`, na.rm = TRUE)

n <- 50

se <- sigma / sqrt(n)

mu
se
```
Although the population distribution of Sleep Duration is not normal, the Central Limit Theorem guarantees that the sampling distribution of the sample mean is approximately normal for a sample size of 50. The mean of this distribution equals the population mean, and its variability is reduced by a factor of $\sqrt{n}$, resulting in a standard error of $\frac{\sigma}{\sqrt{50}}$


Assume $\bar{X} > 7$

```{r}
x_value <- 7  # choose a meaningful threshold

z <- (x_value - mu) / (sigma / sqrt(n))

prob <- 1 - pnorm(z)

z
prob
```
To find the probability that the sample mean exceeds 7 hours, we standardize using the sampling distribution of the mean:

Z = $\frac{\bar{X} - \mu}{\sigma / \sqrt{n}}$

The resulting probability is computed using the standard normal distribution.

# Heading 2: Introduction

## Responsibile Living
Sleep is essential to physical, mental, and emotional health. It plays a key role in maintaining energy levels, regulating emotional responses, and supporting overall bodily function. Research consistently links good health to both the quality and duration of sleep, showing that insufficient or poor sleep can negatively impact well-being over time.

## Research Objectives
The purpose of this analysis is to study patterns in sleep health using both qualitative and quantitative variables from the dataset. The analysis is designed to estimate population characteristics, test statistical claims, compare groups, and examine relationships between variables. This is important because sleep affects health, daily functioning, and overall quality of life. By using confidence intervals and hypothesis tests, the analysis provides evidence-based conclusions rather than simple guesses from raw data.

1. One-sample confidence interval for the population proportion
Variable: Sleep Disorder
* What is a confidence interval for the true proportion of individuals who have no sleep disorder?

2. One-sample confidence interval for the population mean
Variable: Sleep Duration
What is a confidence interval for the true mean sleep duration of individuals in the population?

3. One-sample proportion hypothesis test
Variable: Sleep Disorder
Is the true proportion of individuals with no sleep disorder different from 60%

$H_0: p = 0.60$
$H_A: p \ne 0.60$

4. One-Sample t test
Variable: Sleep Duration
Is the true mean sleep duration different from 7 hours?

$H_0: \mu = 7$
$H_A:\mu\ne7$

5. Two-sample mean independent t test
Variables: Sleep Duration, Gender
Is there a signigicant difference in mean sleep duration between male and female individuals?

$H_0:\mu_M=\mu_F$
$H_A:\mu_M\ne\mu_F$

6. Linear relationship between two quantitative variables
Variables: Sleep Duration, Quality of Sleep
Is there a linear relationship between sleep duration and quality of sleep?

$H_0:\beta_1=0$
$H_A: \beta_1 \ne 0$

## Data Overview

The dataset used in this project is the [Sleep Health and Lifestyle Dataset](https://www.kaggle.com/datasets/uom190346a/sleep-health-and-lifestyle-dataset/data) from Kaggle. It contains information about sleep habits, health indicators, and lifestyle factors for individuals. According to the dataset description, it contains 400 rows and 13 columns, with each row representing one person and each column representing a measured variable.

The key variables used in this project are:

- **Sleep Disorder**: a qualitative variable used for the one-sample confidence interval for a population proportion and the one-sample proportion hypothesis test.
- **Sleep Duration**: a quantitative variable used for the one-sample confidence interval for the population mean and the one-sample t test.
- **Gender**: a qualitative grouping variable used for the two-sample independent t test.
- **Quality of Sleep**: a quantitative variable used with Sleep Duration to examine whether there is a linear relationship between two quantitative variables.

# Heading 3: Exploratory Data Analysis

## Qualitative Variables)

### Sleep Duration)

Summary
```{r}
summary(sleep_df$`Sleep Duration`)
mean(sleep_df$`Sleep Duration`, na.rm = TRUE)
median(sleep_df$`Sleep Duration`, na.rm = TRUE)
sd(sleep_df$`Sleep Duration`, na.rm = TRUE)
IQR(sleep_df$`Sleep Duration`, na.rm = TRUE)
range(sleep_df$`Sleep Duration`, na.rm = TRUE)
```
Sleep Duration Histogram:
```{r}
ggplot(sleep_df, aes(x = `Sleep Duration`)) +
  geom_histogram(binwidth = 0.5, fill = "lightblue", color = "white") +
  labs(
    title = "Distribution of Sleep Duration",
    x = "Sleep Duration",
    y = "Count"
  )
```

The distribution of sleep duration is slightly right-skewed (positively skewed), indicating a small concentration of lower sleep values that extend the lower tail of the distribution. Most observations are clustered between approximately 6.4 and 7.8 hours, with a central tendency around a median of 7.2 hours and a mean of 7.13 hours, suggesting the distribution is fairly balanced overall with only mild skewness.

The histogram also indicates a noticeable concentration of values around 8 hours, forming a secondary peak and suggesting that a portion of the sample reports higher-than-average sleep duration. The relatively small standard deviation (~0.80) reflects low variability, meaning most sleep durations are fairly consistent and close to the mean.

### Quality of Sleep)

Summary
```{r}
summary(sleep_df$`Quality of Sleep`)
mean(sleep_df$`Quality of Sleep`, na.rm = TRUE)
median(sleep_df$`Quality of Sleep`, na.rm = TRUE)
sd(sleep_df$`Quality of Sleep`, na.rm = TRUE)
IQR(sleep_df$`Quality of Sleep`, na.rm = TRUE)
range(sleep_df$`Quality of Sleep`, na.rm = TRUE)
```

Histogram Distribution
```{r}
ggplot(sleep_df, aes(x = `Quality of Sleep`)) +
  geom_histogram(binwidth = 1, fill = "lightgreen", color = "white") +
  labs(
    title = "Distribution of Quality of Sleep",
    x = "Quality of Sleep",
    y = "Count"
  )
```

The distribution of Quality of Sleep is concentrated toward higher values, with most observations falling between 6 and 8, indicating generally good sleep quality across the sample. The median is 7 and the mean is 7.31, suggesting a slight tendency toward higher sleep quality, with only a small difference between the two measures.

The histogram indicates a slight negative (left) skew, driven by a small number of lower scores, particularly the minimum value of 4, which extends the lower tail of the distribution. In contrast, higher values up to 9 are more frequent and form the main cluster of the data.

## Categorical Variables

### Sleep Disorder)

Freq and Rel Freq
```{r}
# Frequency table
table(sleep_df$`Sleep Disorder`)

# Relative frequency table
prop.table(table(sleep_df$`Sleep Disorder`))
```
Bar Graph:
```{r}
ggplot(sleep_df, aes(x = `Sleep Disorder`)) +
  geom_bar(fill = "violet") +
  labs(
    title = "Sleep Disorder Distribution",
    x = "Sleep Disorder",
    y = "Count"
  )
```

The distribution of sleep disorder status shows that the majority of participants (58.6%) fall into the `None` category, indicating that most individuals in the sample do not report a diagnosed sleep disorder. This dominance of the `None` group is the most prominent feature of the data.

In contrast, Insomnia (20.6%) and Sleep Apnea (20.9%) are present at much lower and nearly equal proportions. This suggests that, among those with a reported sleep disorder, the sample is fairly evenly split between the two conditions.

### Gender)

Freq
```{r}
table(sleep_df$Gender)
prop.table(table(sleep_df$Gender))
```
Bar Graph
```{r}
ggplot(sleep_df, aes(x = Gender)) +
  geom_bar(fill = "lightpink") +
  labs(
    title = "Gender Distribution",
    x = "Gender",
    y = "Count"
  )
```
The distribution of gender is nearly even, with 189 males (50.5%) and 185 females (49.5%). This indicates a well-balanced sample with no meaningful gender imbalance.

The bar graph reflects this symmetry clearly, with both categories appearing almost equal in height. Overall, gender is evenly represented in the dataset.

# Heading 4: One Sample: Confidence Interval and Hypothesis Testing

## Section 4.0: Population and sample

The population of interest is the broader group of individuals whose sleep pattern are represented by this dataset.

The sample is the set of individuals recorded in the Sleep Health and Lifestyle Dataset that were used in this analysis.

## Section 4.1: Quanitative Variable

### Check assumptions

- The variable Sleep Duration is quantitative
- The sample size is large, with more than 30 observations
- The observations are assumed to be independent because each row represents a different individual
- The assumptions appear reasonable for this analysis

The quantitative variable used in this section is Sleep Duration, measured in hours of sleep per individual.

This is a quantitative numerical variable.

The parameter of interest is the population mean sleep duration, denoted by $\mu$.

The statistic is the sample mean sleep duration, denoted by \(\bar{x}\).

The research question is: Is the true mean sleep duration different from 7 hours?

$H_0: \mu = 7$
$H_A: \mu \ne 7$

```{r}
sleep_duration <- sleep_df$`Sleep Duration`

xbar <- mean(sleep_duration)
xbar

# 99% confidence interval for pop mean
t.test(sleep_duration, conf.level = 0.99)

# One-sample t test for mu = 7
t.test(sleep_duration, mu = 7, alternative = "two.sided", conf.level = 0.99)
```

The point estimate for the population mean sleep duration is the same mean $\bar{x}$

We are 99% confident that the true population mean sleep duration lies within 7.02 - 7.23 hours. This means that if we were to repeatedly take samples and construct confidence intervals in the same way, approximately 99% of those intervals would contain the true mean sleep duration.

Since the p-value = 0.00144, which is less than 0.01, we reject the null hypothesis.
There is sufficient statistical evidence to conclude that the true mean sleep duration is different from 7 hours.

## 4.2: Qualitative variable

### Check Assumptions

- The variable `Sleep Disorder` is categorical.
- The observations are assumed to be independent because each row represents a different individual.
- The number of successes is at least 10.
- The number of failures is at least 10.
- The success-failure condition is met.
- The assumptions appear reasonable for this analysis.

The qualitative variable used in this section is Sleep Disorder.

For this analysis, a success is defined as an individual having no sleep disorder, that is, `Sleep Disorder = "None"`.

This is a categorical variable.

The parameter of interest is the population proportion of individuals with no sleep disorder, denoted by $p$.

The statistic is the sample proportion of individuals with no sleep disorder, denoted by $\hat{p}$.

The research question is: Is the true proportion of individuals with no sleep disorder different from 60%?

$H_0: p = 0.60$
$H_A: p \ne 0.60$

```{r}
x_success <- sum(sleep_df$`Sleep Disorder` == "None")
n_total <- sum(!is.na(sleep_df$`Sleep Disorder`))
phat <- x_success / n_total

# Point estimate
phat

# 98% confidence interval for pop proportion
prop.test(x_success, n_total, conf.level = 0.98)

# One-sample proportion test for p = 0.60
prop.test(x_success, n_total, p = 0.60, alternative = "two.sided")
```
We are 98% confident that the true population proportion of individuals with no sleep disorder lies within 54.36% - 63%. This suggests that the actual proportion in the population is likely within this range.

Since the p-value = 0.605, which is greater than 0.02, we do not reject the null hypothesis.
There is insufficient statistical evidence to conclude that the true proportion of individuals with no sleep disorder is different from 60%.

# Heading 5: Two Samples T-Test - Independent Sampling Method

### Check Assumptions

- The response variable `Sleep Duration` is quantitative.
- The explanatory variable `Gender` is categorical with two independent groups: male and female.
- The observations are assumed to be independent because each row represents a different individual.
- Both groups have sample sizes greater than 30.
- The assumptions appear reasonable for this analysis.

Research Question:
  "Is there a difference in sleep duration based on gender?"

Population:
  All Individuals who identified as `Male`
  All Individuals who identified as `Female`
  
Samples:
  Sleep Duration in Men
  Sleep Duration in Women
  
Variables:
  `Gender`: Gender is a categorical variable representing the `X` value. Made up of 2 independent groups, individuals identifying as Male, and individuals identifying as Female.
  `Sleep Duration`: Sleep Duration is a quantitative variable representing the Y value. It's a continuous numerical variable for average number of hours of sleep a night.
  
Parameters:
  $\mu_m$ = Avg. Sleep Duration for Males
  $\mu_f$ = Avg. Sleep Duration for Females
  
Statistics:
  $\bar{x}_1$ = Sample Mean Sleep Duration for Males
  $\bar{x}_2$ = Sample Mean Sleep Duration for Females
  
Hypothesis:  
  
$H_0: \mu_m - \mu_f = 0$

$H_A: \mu_m - \mu_f \neq 0$

```{r}
# Two-sample independent t-test
t_test_result <- t.test(`Sleep Duration` ~ Gender, data = sleep_df,
                        alternative = "two.sided",
                        var.equal = FALSE)

t_test_result
```
A Welch two-sample t-test was conducted to compare mean sleep duration between females and males. The results showed a statistically significant difference in mean sleep duration between the two groups, t(349.38) = 2.36, p = 0.019. Since the p-value is less than 0.05, we reject the null hypothesis that the population means are equal.

The 95% confidence interval for the difference in means was (0.032, 0.354), which does not include 0. This supports the conclusion that there is a significant difference in average sleep duration between genders.

The sample estimates indicate that females (M = 7.23) report slightly higher sleep duration than males (M = 7.04).

# Heading 6: Sample Linear Regression

- Both variables, `Sleep Duration` and `Quality of Sleep`, are quantitative.
- A scatterplot should be used to check that the relationship between the two variables is approximately linear.
- The observations are assumed to be independent because each row represents a different individual.
- The assumptions appear reasonable for this analysis.

Research Question:
  Is there a linear relationship between Sleep Duration and Quality of Sleep?

X: Sleep Duration: A Quantitative Value representative of hours of sleep.
Y: Quality of Sleep: A Quantitative Value representative of a 0-10 scale of Sleep Quality.

Hypothesis:
$H_0: \beta_1 = 0$
$H_A: \beta_1 \neq 0$

```{r}
model <- lm(sleep_df$`Quality of Sleep` ~ sleep_df$`Sleep Duration`)

summary(model)

ggplot(sleep_df, aes(x = `Sleep Duration`, y = `Quality of Sleep`)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Sleep Duration vs Sleep Quality",
    x = "Sleep Duration (hours)",
    y = "Quality of Sleep"
  ) +
  theme_minimal()
```
The p-value associated with the slope coefficient is less than 0.001. Since this value is less than 0.05, we reject the null hypothesis that$\beta_{1} = 0$

This indicates that there is a statistically significant linear relationship between sleep duration and quality of sleep.

The model’s $R^{2}$ value of approximately 0.78 indicates that about 78% of the variability in sleep quality is explained by sleep duration.

# Heading 7: Report

Overall, the analysis suggests that sleep duration, sleep quality, gender, and sleep disorder status are meaningful variables for studying well-being and responsible living. Most individuals in the dataset reported no sleep disorder, though insomnia and sleep apnea were still present in a substantial portion of the sample. The average sleep duration was close to 7 hours, and the hypothesis test examined whether the true mean differed from that benchmark.

The two-sample t-test showed a statistically significant difference in average sleep duration between males and females, with females reporting slightly higher sleep duration on average. The regression analysis also showed a strong positive relationship between sleep duration and quality of sleep, meaning individuals who slept longer tended to report better sleep quality.

The linear regression analysis showed that 78% of the variation in sleep quality ratings can be explained by the relationship between sleep duration and sleep quality. This suggests a strong linear association between how long a person sleeps and how they rate the quality of their sleep.

These results connect to responsible living because sleep is an important part of physical health, mental health, and daily functioning. The analysis supports the idea that lifestyle and health patterns can be studied statistically, allowing conclusions to be based on evidence rather than assumptions.
