---
title: "GarnerChildersMuhimpundu_FinalProject"
author: "Seth Garner, Gavin Childers, Julie Muhimpundu"
date: "2026-04-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..")
library(tidyverse)
```

```{r}
sleep_df <- read_csv("sleep_data.csv")
sleep_df
```

# Heading 1: Probability Distributions

# 1.1) Qualitative Data

Sleep Disorder
Success == "None"
Fail == "Sleep Apnea" or "Insomnia"

```{r}
# Population proportion (p)
p <- mean(sleep_df$`Sleep Disorder` == "None", na.rm = TRUE)
p

n <- 50

# Mean (mu)
mu <- n * p

# Standard deviation (sigma)
sigma <- sqrt(n * p * (1 - p))

mu
sigma
```
The random variable X, representing the number of individuals with no sleep disorder in a sample of size 50, follows a binomial distribution because:
i) The Number of Trials are Fixed (n = 50)
ii) Each trial has two outcomes (Success, Fail)
iii) The probability of success (p) is constant
iv) Observations are Independent

```{r}
np <- n * p
nq <- n * (1 - p)
np
nq

z <- (mu + 0.5 - mu) / sigma
z

prob <- 1 - pnorm(z)
prob
```
Since np & n(1-p) > 10, the binomial distribution is approximately normal. A continuity correction of 0.5 is applied when converting from the discrete binomial distribution to the continuous normal distribution. The resulting probability represents the likelihood that the number of individuals with no sleep disorder in a sample of 50 exceeds the expected value.

```{r}
# Population proportion
p <- mean(sleep_df$`Sleep Disorder` == "None", na.rm = TRUE)

# Sample size
n <- 50

# Mean of sampling distribution
phat_mean <- p

# Standard error of sampling distribution
phat_se <- sqrt(p * (1 - p) / n)

phat_mean
phat_se
```
The mean of the sampling distribution equals the population proportion p, confirming that $\hat{p} $is an unbiased estimator. The standard error measures the typical variation of the sample proportion from sample to sample, and it decreases as sample size increases.

# 1.2) Quanitative Data

Sleep Duration

The quantitative variable selected is `Sleep Duration`, which measures the number of hours of sleep per individual.

```{r}
ggplot(sleep_df, aes(x = `Sleep Duration`)) +
  geom_histogram(bins = 30, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Sleep Duration",
       x = "Sleep Duration",
       y = "Frequency")
```
The histogram of Sleep Duration does not show a clear bell-shaped or symmetric distribution. Instead, the data appears somewhat unevenly distributed, with mild clustering and no strong single central peak. This suggests that the variable is not approximately normally distributed.

Because the shape is not clearly symmetric and does not follow the typical Gaussian pattern, we conclude that the population distribution of Sleep Duration is not normal.

The sampling distribution of the sample mean can still be approximately normal due to the Central Limit Theorem.


Sample Size: n = 50

X = Sleep Duration
$\bar{X}$ = Sample Mean of Sleep Duration
n = 50

```{r}
mu <- mean(sleep_df$`Sleep Duration`, na.rm = TRUE)
sigma <- sd(sleep_df$`Sleep Duration`, na.rm = TRUE)

n <- 50

se <- sigma / sqrt(n)

mu
se
```
Although the population distribution of Sleep Duration is not normal, the Central Limit Theorem guarantees that the sampling distribution of the sample mean is approximately normal for a sample size of 50. The mean of this distribution equals the population mean, and its variability is reduced by a factor of $\sqrt{n}$, resulting in a standard error of $\frac{\sigma}{\sqrt{50}}$


Assume $\bar{X} > 7$

```{r}
x_value <- 7  # choose a meaningful threshold

z <- (x_value - mu) / (sigma / sqrt(n))

prob <- 1 - pnorm(z)

z
prob
```
To find the probability that the sample mean exceeds 7 hours, we standardize using the sampling distribution of the mean:

Z = $\frac{\bar{X} - \mu}{\sigma / \sqrt{n}}$

The resulting probability is computed using the standard normal distribution.

# Heading 2: Introduction

# Research Objectives
The purpose of this analysis is to study patterns in sleep health using both qualitative and quantitative variables from the dataset. The analysis is designed to estimate population characteristics, test statistical claims, compare groups, and examine relationships between variables. This is important because sleep affects health, daily functioning, and overall quality of life. By using confidence intervals and hypothesis tests, the analysis provides evidence-based conclusions rather than simple guesses from raw data.

1. One-sample confidence interval for the population proportion
Variable: Sleep Disorder
* What is a confidence interval for the true proportion of individuals who have no sleep disorder?

2. One-sample confidence interval for the population mean
Variable: Sleep Duration
What is a confidence interval for the true mean sleep duration of individuals in the population?

3. One-sample proportion hypothesis test
Variable: Sleep Disorder
Is the true proportion of individuals with no sleep disorder different from 60%

$H_0: p = 0.60$
$H_A: p \ne 0.60$

4. One-Sample t test
Variable: Sleep Duration
Is the true mean sleep duration different from 7 hours?

$H_0: \mu = 7$
$H_A:\mu\ne7$

5. Two-sample mean independent t test
Variables: Sleep Duration, Gender
Is there a signigicant difference in mean sleep duration between male and female individuals?

$H_0:\mu_M=\mu_F$
$H_A:\mu_M\ne\mu_F$

6. Linear relationship between two quantitative variables
Variables: Sleep Duration, Quality of Sleep
Is there a linear relationship between sleep duration and quality of sleep?

$H_0:\beta_1=0$
$H_A:\beta_1!=0$

# Data Overview

The dataset used in this project is the Sleep Health and Lifestyle Dataset. It contains information about sleep habits, health indicators, and lifestyle factors for individuals. According to the dataset description, it contains 400 rows and 13 columns, with each row representing one person and each column representing a measured variable.

The key variables used in this project are:

- **Sleep Disorder**: a qualitative variable used for the one-sample confidence interval for a population proportion and the one-sample proportion hypothesis test.
- **Sleep Duration**: a quantitative variable used for the one-sample confidence interval for the population mean and the one-sample t test.
- **Gender**: a qualitative grouping variable used for the two-sample independent t test.
- **Quality of Sleep**: a quantitative variable used with Sleep Duration to examine whether there is a linear relationship between two quantitative variables.

# Heading 3: Exploratory Data Analysis

# Heading 4: One Sample: Confidience Interval and Hypothesis Testing

## Section 4.0: Population and sample

The population of interest is the broader group of individuals whose sleep pattern are represented by this dataset.

The sample is the set of individuals recorded in the Sleep Health and Lifestyle Dataset that were used in this analysis.

## Section 4.1: Quanitative Variable

The quantitative variable used in this section is Sleep Duration, measured in hours of sleep per individual.

This is a quantitative numerical variable.

The parameter of interest is the population mean sleep duration, denoted by $\mu$.

The statistic is the sample mean sleep duration, denoted by \(\bar{x}\).

The research question is: Is the true mean sleep duration different from 7 hours?

$H_0: \mu = 7$
$H_A:\mu!=7$

```{r}
sleep_duration <- sleep_df$`Sleep Duration`

xbar <- mean(sleep_duration)
xbar

# 99% confidence interval for pop mean
t.test(sleep_duration, conf.level = 0.99)

# One-sample t test for mu = 7
t.test(sleep_duration, mu = 7, alternative = "two.sided", conf.level = 0.99)
```

The point estimate for the population mean sleep duration is the same mean $\bar{x}$

A 99% confidence interval was constructed for the true population mean sleep duration.

A one-sample t test was also conducted to test whether the true mean sleep duration differes from 7 hours

## 4.2: Qualitative variable

The qualitative variable used in this section is Sleep Disorder.

For this analysis, a success is defined as an individual having no sleep disorder, that is, `Sleep Disorder = "None"`.

This is a categorical variable.

The parameter of interest is the population proportion of individuals with no sleep disorder, denoted by $p$.

The statistic is the sample proportion of individuals with no sleep disorder, denoted by $\hat{p}$.

The research question is: Is the true proportion of individuals with no sleep disorder different from 60%?

$H_0: p = 0.60$
$H_A: p \ne 0.60$

```{r}
x_success <- sum(sleep_df$`Sleep Disorder` == "None")
n_total <- sum(!is.na(sleep_df$`Sleep Disorder`))
phat <- x_success / n_total

# Point estimate
phat

# 98% confidence interval for pop proportion
prop.test(x_success, n_total, conf.level = 0.98)

# One-sample proportion test for p = 0.60
prop.test(x_success, n_total, p = 0.60, alternative = "two.sided")
```

# Heading 5: Two Samples T-Test - Independent Sampling Method

# Heading 6: Sample Linear Regression

# Heading 7: Report